---
title: "Military Coups"
output: html_notebook
---


```{r,echo = FALSE}
rm(list = ls())
library(tidyverse)
theme_set(theme_minimal())
```

```{r, echo=FALSE}
library(broom)
library(gt)
```
(a) We read the data set from coups.txt file and is stored in the variable data. Convert political liberalization levels into a factor, we treat it as a categorical variable. We also inspect the variable data 

```{r, echo=FALSE}
data <- read.table("coups.txt", header = TRUE, sep = "", stringsAsFactors = FALSE)
data$pollib <- as.factor(data$pollib)
glimpse(data)
```
We perform a Poisson regression analysis on the dataset and display a summary of the model
```{r, echo=FALSE}
# Perform Poisson regression
poisson_model <- glm(miltcoup ~ oligarchy + pollib + parties + pctvote + popn + size + numelec + numregim, 
                     data = data, 
                     family = poisson)

# View model summary
summary(poisson_model)

```
The Poisson regression model shows that oligarchy, pollib2, and parties are significant predictors of military coups (p < 0.05). Specifically, more years of military oligarchy and more legal political parties increase coups, while higher political liberalization (pollib2) reduces coups. Non-significant predictors (pctvote, popn, size, etc.) can be considered for removal to simplify the model.

(b)
We start reducing the explanatory variables in the model by using the step down approach
```{r, echo=FALSE}
reduced_model_1 <- glm(miltcoup ~ oligarchy + pollib + parties + pctvote + popn + size + numregim, 
                       data = data, 
                       family = poisson(link = "log"))
summary(reduced_model_1)
  
```
Number of legislative and presidential elections  has a p-value of $0.670$ so we remove it from the model and re-evaluate the mode we keep on doing this till we can no longer eliminate any other variable
Iteratively we remove the explanatory variables (in order) number of regime types, area in 1000 square km, population in millions in 1989.

```{r, echo=FALSE}
reduced_model_4 <- glm(miltcoup ~ oligarchy + pollib + parties + pctvote, 
                       data = data, 
                       family = poisson(link = "log"))
summary(reduced_model_4)
```
The next variable to remove is a categorical variable.We recode the pollib variable to combine levels 0 and 1 into a single group (0) and treat level 2 as a separate group (1). This simplifies the model by focusing on the significant difference between pollib = 2 (higher political liberalization) and the combined group of pollib = 0 and 1 (lower liberalization), while reducing complexity and improving interpretability.
```{r, echo=FALSE}
data$pollib_recoded <- ifelse(data$pollib == 2, 1, 0)

recoded_model <- glm(miltcoup ~ oligarchy + pollib_recoded + parties + pctvote, 
                   data = data, 
                   family = poisson(link = "log"))
summary(recoded_model)
```
The we remove the  variable percent voting in last election since it is insignificant p-value= $0.255$ 
```{r, echo=FALSE}
final_model <- glm(miltcoup ~ oligarchy + pollib_recoded + parties, 
                           data = data, 
                           family = poisson(link = "log"))
summary(final_model)
```
The final model simplifies the initial model by retaining only significant predictors (oligarchy, pollib_recoded, parties). Both models show that oligarchy and parties increase coups, while pollib_recoded (or pollib2 in the initial model) reduces coups.The final model has a lower AIC ($106.63$ vs. $113.06$), indicating a better balance of fit and simplicity. Removing non-significant predictors improves interpretability without sacrificing key insights.

(c)We find the expected number of Coups for different Political Liberalization levels we consider pollibs 0 and 1 to not have any significant difference
```{r, echo=FALSE}

avg_oligarchy <- mean(data$oligarchy)
avg_parties <- mean(data$parties)

cat("Average oligarchy:", avg_oligarchy, "\n")
cat("Average parties:", avg_parties, "\n")
```
```{r, echo=FALSE}
predict_coups <- function(oligarchy, parties, pollib_recoded) {

  linear_predictor <- coefficients["(Intercept)"] + 
                      coefficients["oligarchy"] * oligarchy + 
                      coefficients["pollib_recoded"] * pollib_recoded + 
                      coefficients["parties"] * parties
  

  predicted_coups <- exp(linear_predictor)
  
  return(predicted_coups)
}
```
```{r, echo=FALSE}
coefficients <- coef(final_model)
predicted_coups_0 <- predict_coups(oligarchy = avg_oligarchy, 
                                   parties = avg_parties, 
                                   pollib_recoded = 0)

predicted_coups_1 <- predict_coups(oligarchy = avg_oligarchy, 
                                   parties = avg_parties, 
                                   pollib_recoded = 1)

cat("Predicted number of coups for None or Limited Civil Rights):", predicted_coups_0, "\n")
cat("Predicted number of coups for Full Civil Rights:", predicted_coups_1, "\n")
```
Countries with full civil rights (pollib_recoded = 1) are predicted to have fewer military coups $(0.969)$ compared to those with none or limited civil rights (pollib_recoded = 0, $1.985$ coups). oligarchy and parties increase coups, while pollib_recoded significantly reduces them.




